Project Diary

### 2014-05-10 thesis/price_models.tex
Revision following Sven's advice.

The justification of modeling return series as independent,
identically distributed Gaussian variates comes from imagining the
price process $S(t)$ as a Brownian motion whose increment $\sigma
dw_t$ at each time step is independent and scales as $\sigma
\sqrt{dt}$. Then, by adding a drift term $\mu dt$ that represents some
deterministic trend in the price process, one can express the
price $S(t)$ as a stochastic differential equation:
\begin{eqnarray*}
  dS &=& S\mu dt + S\sigma dw_t \\
  d(\ln S) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$. Therefore, in a discrete-time
model, where the length of each time step $\Delta t$ is fixed, the
return $r_t = \ln S(t + \Delta t) - \ln S(t)$ is assumed to be
Gaussian distributed and have mean and variance that are functions of
$\Delta t$.

The picture depicted above is of course overly simplified, and as we
discussed in \S\ref{sec:StylizedFacts}, the distribution of returns is
not really Gaussian. Nevertheless, the assumption of Gaussian
distributed returns underlies such important theories as Black and
Scholes theory of option pricing. Thus, albeit hugely inaccurate, the
Gaussian distribution as a returns model shall not be forgotten. In
the next few sections, we discuss some more realistic returns models.

\item{\bf GARCH models}
``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A GARCH(p, q) model is defined by the following
equation system:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu + \epsilon_t \\
    \sigma_t^2 &= \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
    \sum_{i=1}^p \beta_i \sigma_{t-i}^2
  \end{aligned}
\end{equation}
where $\epsilon_t$ is termed the innovation of the return $r_t$ and is
a random variable of a prescribed distribution, the most common ones
being Gaussian, Student's t, Asymmetric Student's t, etc; $\sigma_t^2$
is the variance of $\epsilon_t$ \cite{Bollerslev86}. The
autocorrelation function of $\epsilon_t^2$ is given by
(c.f. \cite{Bollerslev87})
\begin{eqnarray*}
\rho_n &=& \sum_{i=1} ^{p \vee q} (\alpha_i + \beta_i) \rho_{n-i}
\text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $\max(p, q)$.

\item {\bf Stochastic Volatility Models}\label{sec:SLV_model}
For the purpose of intraday returns, we specify the stochastic
volatility model as
\begin{eqnarray}
   r_{t, t-h} &=& \ln p_{t} - \ln p_{t-h} \nonumber \\
   r_{t, t-h} &=& \mu + \sigma_{t, t-h} b_t \label{eq:SLV_spec}
\end{eqnarray}
where $p_t$ is the price of the asset at time $t$; $b_t \sim N(0,
1)$; $r_{t, t-h}$ is the return in the time interval $[t-h, t]$.
For simplicity, in the rest of this chapter we shall just write
$t$ for the subscript ``t, t-h'', since the time interval $h$ is fixed
for each time series and is a known constant.

Andersen and Bollerslev et al proved the following in
\cite{Andersen03} (theorem 2):
\begin{equation}
  \label{eq:normal_r}
  r_t|\mathcal{F}_{t-h} \sim N(\int_{0}^h \mu_{t-h+s} ds, \int_{0}^h
  \sigma_{t-h+s}^2 ds)
\end{equation}
In plain words, given all the information up to time $t-h$, the
distribution of $r_t$ is Gaussian with integrated mean and variance.
The variance of this conditional distribution, $\int_{0}^h
\sigma_{t-h+s}^2 ds$, can be approximated by \cite{Protter05,
  Andersen03}:
\begin{equation}
  \label{eq:rv_def}
  \int_{0}^h \sigma_{t-h+s}^2 ds = \sum_{k=1}^n \left(
    p_{t-h+kh/n} - p_{t-h+(k-1)h/n} \right)^2
\end{equation}
where $n$ is a chosen constant. The square root of the right hand side
of equation \ref{eq:rv_def} is the realized volatility, call it
$\hat{\sigma}_t$.

@@

To compare GARCH and stochastic log-volatility models at the face of
intraday returns, we study the {\it Nordea Bank} 15-minute returns
during the period 2012/01/16 - 2012/04/20 in \S\ref{chp:nordea_15min},
and {\it Volvo B} 30-minute returns during 2013/10/10 - 2014/04/04.

\section{Case Study: Nordea in 1st quarter 2012}
\label{chp:nordea_15min}
In this section we investigate the Nordea 15-minute returns sampled
during the period 2012/01/16 - 2012/04/20. In total, these amount to
2022 returns. We use the first 80\% (1617) for model estimation and  the
remaining 20\% (405) for comparing with model forcasts. In section
\ref{sec:nordea_15min_garch} we study the series with a GARCH model 
and in section \ref{sec:nordea_15min_arima} we study it with a
stochastic volatility model.

\subsection{GARCH Model}\label{sec:nordea_15min_garch}
First of all, to check for GARCH effects, we plot the auto-correlation
function (ACF) of the squared returns. This is shown in figure
\ref{fig:nordea_15min_acf}. At the absence of GARCH effects, the ACF
is expected to have an asymptotic Gaussian distribution with mean 0
and variance 1/T \cite{Bollerslev86, Bollerslev87}, which is clearly
not the case in figure \ref{fig:nordea_15min_acf} --- the first 5
autocorrelations have comparable sizes and do not fall off as in a
Gaussian scheme. Moreover, figure \ref{fig:nordea_15min_vlt_acf} shows
even more clearly that the conditional variances of the series are
correlated. These observations suggest a GARCH(p, q) model can be
appropriate.

Starting with a GARCH(1,1) model and taking advantage of the knowledge
that the log-volatility $\ln \sigma_t$ has seasonality $s=33$, we fit
to the return series a GARCH(33, 33) model, limiting to lags 1 and 33
for both ARCH and GARCH parameters.
\begin{eqnarray*}
  r_t &=& \mu + €_t \\
  €_t &=& \sigma_t z_t \\
  \sigma^2_t &=& \alpha_0 + \alpha_1 €^2_{t-1} + \alpha_s €^2_{t-s} +
  \beta_1 \sigma^2_{t-1} + \beta_s \sigma^2_{t-s}
\end{eqnarray*}
Via maximum likelihood estimation, parameter values listed in table
\ref{tab:nordea_15min_garch} are obtained.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    Parameter & $\alpha_0$ & $\alpha_1$ & $\alpha_s$ & $\beta_1$ &
    $\beta_s$ \\
    \hline
    Value & $4.7833 \times 10^{-7}$ & 0.1600 & 0.0667 & 0.6846 &
    0.0342 \\
    \hline
  \end{tabular}
  \caption{\small \it GARCH model parameters}
  \label{tab:nordea_15min_garch}
\end{table}

@@

The above seasonal moving average model has the following
autocovariance structure \cite{BoxJenkins94}:
\begin{eqnarray*}
  \gamma_0 &=& \sigma_y^2 (1 + \theta^2)(1 + \Theta^2) \\
  \gamma_1 &=& -\sigma_y^2\theta(1 + \Theta^2) \\
  \gamma_s &=& -\sigma_y^2\Theta(1 + \theta^2) \\
  \gamma_{s+1} &=& \gamma_{s-1}\;=\;\sigma_y^2\theta\Theta
\end{eqnarray*}
These equations together with the measured autocorrelations make
possible an initial estimate of the parameters $\theta$ and $\Theta$:
\begin{eqnarray*}
  {\rho_{s+1}/\rho_s} &=& {\gamma_{s+1}/\gamma_s} \;=\; -{\theta \over
    1 + \theta^2} \\
  {\rho_{s+1}/\rho_1} &=& {\gamma_{s+1}/\gamma_1} \;=\; -{\Theta \over
    1 + \Theta^2} \\
\end{eqnarray*}
Substituting in the measured values shown in table
\ref{tab:nordea_15min_w_acf},
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    $\rho_1$ & $\rho_{s-1}$ & $\rho_s$ & $\rho_{s+1}$ \\
    \hline
    -0.4703 &  0.2053 & -0.4564 &  0.2212 \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min $w_t$ autocorrelations}
  \label{tab:nordea_15min_w_acf}
\end{table}
we get
\begin{eqnarray*}
  \theta &=& 0.6890 \\
  \Theta &=& 0.6378
\end{eqnarray*}
Among the two roots of each of the 2nd order equations in the above,
we have chosen the one in the range $(-1, 1)$ so as to ensure
invertibility of the model \cite{BoxJenkins94}.

@@

The first 4 moments of the Johnson Su distribution are expressible
in closed form in $\gamma, \delta, \lambda, \xi$ \cite{Shang2004}:
\begin{eqnarray*}
  w &=& \exp{1 \over \delta^2} \\
  \Omega &=& {\gamma \over \delta} \\
  \text{E}(y) &=& -w^{1/2} \lambda \sinh\Omega + \xi\\
  \text{std}(y) &=& \lambda \left[{1 \over 2}(w-1)(w\cosh 2\Omega +
    1)\right]^{1/2} \\
  \text{skewness}(y) &=& {
    \sqrt{(1/2)w(w-1)} [w(w+2)\sinh 3\Omega + 3\sinh\Omega]
    \over
    (w\cosh 2\Omega + 1)^{3/2}} \\
  \text{kurtosis}(y) &=& {
    w^2(w^4 + 2w^3 + 3w^2 - 3)\cosh 4\Omega + 4w^2 (w+2) \cosh 2\Omega
    + 3(2w+1) \over
    2(w\cosh 2\Omega + 1)^2 }
\end{eqnarray*}
By matching the theoretical expressions of the moments with their
measured values, and taking help from published tables
\cite{Johnson1965}, one can solve for the parameters $\gamma, \delta,
\lambda, \xi$.

@@

The eventual MLE is done in Matlab with the ``active set''
algorithm. The initial as well as the final estimation results are
listed in table \ref{tab:nordea_15min_js_param}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|c|}
    \hline
    & $\gamma$ & $\delta$ & $\lambda$ & $\xi$ & $\theta$ & $\Theta$ \\
    \hline
    initial estimate & 0.1476 & 1.5121 & 0.3633 & 0.0454 & 0.6890 &
    0.6378 \\
    \hline
    MLE estimate & 0.1319 & 1.5266 & 0.3735 & 0.0410 & 0.6639 & 0.6025
    \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min estimation results}
  \label{tab:nordea_15min_js_param}
\end{table}

@@

\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.55, clip=true, trim=27 287 0
    154]{../pics/nordea_2012.pdf}
  \caption{\small \it Blue: SV forecasts; Green: GARCH forecasts; Red:
    sample mean forecasts. Left: cumulative distribution function;
    Right: complementary distribution function.}
  \label{fig:nordea_2012}
\end{figure}

Another measure of the forecasts' quality can be the percentage of good
forecasts, where the criterion of ``good'' is defined, respectively,
as deviating no more than 1\%, 5\%, or 10\% from the measured value of
the realized volatility. Table \ref{tab:nordea_2012_good} shows the
respective percentage of the 3 kinds of forecasts. Again in this table
it is seen that the SV model gives more accurate forecasts than does
GARCH.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${\ln \sigma^F_t - \ln \hat{\sigma}_t \over |\ln \hat{\sigma}_t|}$
    & SV & GARCH & sample mean \\
    \hline
    1\% & 0.2173 & 0.1383 & 0.1136 \\
    \hline
    5\% & 0.7457 & 0.7062 & 0.5333 \\
    \hline
    10\% & 0.9728 & 0.9383 & 0.8914 \\
    \hline
  \end{tabular}
  \caption{\small \it The percentage of ``good'' forecasts when the
    criterion of being good is deviating no more than 1\%, 5\% or
    10\%.}
  \label{tab:nordea_2012_good}
\end{table}

\section{Case Study: Volvo in 2013/14}
\label{sec:volvo}
In this section we model the log-volatility of {\it Volvo B} 30-minute
returns during the period 2013/10/10 - 2014/04/04. This series
contains 1884 log-volatilities computed using 2-minute returns in each
30-minute interval. Among them we use the first 1507 for model
estimation and the last 377 for forecast and model verification.

The left plot of figure \ref{fig:volvo_inno_and_lv_acf} shows the
distribution of $(r_t - \E(r_t))/\sigma_t$. We observe in the figure a
nice Gaussian variate, so we can be sure that the sum of squared
2-minute returns makes a good approximation to the variance of
30-minute returns in this particular case.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=0 256 0
    151]{../pics/volvo_inno_and_lv_acf.pdf}
    \label{fig:volvo_inno_and_lv_acf}
  \caption{\small \it Left: probability plot of $(r_t -
    \E(r_t))/\sigma_t$; Right: auto-correlations of $\ln \sigma_t$}
\end{figure}

@@

\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=10 282 0
    243]{../pics/volvo_slv_garch_cmp.pdf}
    \label{fig:volvo_slv_garch_cmp}
  \caption{\small \it Comparison of the forecast from SV, GARCH,
    and sample mean. Blue: SV forecast; Green: GARCH forecast; Red:
    Sample mean forecast.}
\end{figure}
Figure \ref{fig:volvo_slv_garch_cmp} shows, as in the previous case of
Nordea Bank 15-minute returns, the SV model performs the best, GARCH
the second, and the sample mean the worst. However, when it comes to
over-estimates, the sample mean appears to be the best estimator,
while SV excels over GARCH. But a check of $\E(\ln\sigma_t)$ over
the sample for estimation (1507 data points) and over the sample for
comparison (377 data points) reveals that the first sample has mean
-6.3127 while the second has -6.2072. This increment explains the low
probability of over-estimation when using the first sample mean as
forecast.

Table \ref{tab:volvo_good_percentage} compares the fraction of
``good'' estimates as measured by ${|\ln\sigma^F_t -
  \ln\hat{\sigma}_t| \over |\ln \hat{\sigma}_t|}$ being less than 1\%,
5\% and 10\%.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ &
    SV & GARCH & Sample Mean \\
    \hline
    1\% & 0.2175 & 0.1220 & 0.1406 \\
    \hline
    5\% & 0.7162 & 0.6154 & 0.6631 \\
    \hline
    10\% & 0.9231 & 0.8806 & 0.8992 \\
    \hline
  \end{tabular}
  \caption{\small \it Fraction of ``good'' forecasts as defined by
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ being less than 1\%, 5\% and 10\%.}
  \label{tab:volvo_good_percentage}
\end{table}
We see from the table that the SV model consistently excels over
the other two alternatives.

>>>>

The justification of modeling return series as independent,
identically distributed Gaussian variates comes from imagining the
price process $S(t)$ as a geometric Brownian motion whose increment
$\sigma dw_t$ at each time step is independent and scales as $\sigma
\sqrt{dt}$. Then, by adding a drift term $\mu dt$ that represents some
deterministic trend in the price process, one can express the
price $S(t)$ as a stochastic differential equation:
\begin{eqnarray*}
  dS &=& S\mu dt + S\sigma dw_t \\
  d(\ln S) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$. Therefore, in a discrete-time
model, where the length of each time step $\Delta t$ is fixed, the
return $r_t = \ln S(t + \Delta t) - \ln S(t)$ is assumed to be
Gaussian distributed and have mean and variance that are functions of
$\Delta t$.

The picture depicted above is of course overly simplified, and the
distribution of returns is not really Gaussian. Nevertheless, the
assumption of Gaussian distributed returns underlies such important
theories as Black and Scholes theory of option pricing. Thus, albeit
inaccurate in describing very large returns, the Gaussian distribution
as a returns model shall not be forgotten. In the next few sections,
we discuss some more realistic returns models.

\item{\bf GARCH models}

``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A GARCH(p, q) model is defined by the following
equation system \cite{Bollerslev86}:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu_t + \epsilon_t \\
  \end{aligned}
\end{equation}
where $p$ and $q$ are constant integers; $\mu(t)$ is the mean process
and considered a small constant for intraday returns. $\epsilon_t$ is
called the innovation of the return and is a variate whose conditional
distribution \footnotemark is Gaussian and has variance $\sigma^2_t$:
\begin{eqnarray*}
  \sigma_t^2 &=& \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
  \sum_{i=1}^p \beta_i \sigma_{t-i}^2
\end{eqnarray*}
\footnotetext{
By conditional distribution we mean the distribution conditional on
the variate's history: The conditional distribution at time $t$ means
the distribution conditional on the history up to time $t-1$
in the discrete-time setting.
}

It can be shown that the autocorrelation function of $\epsilon_t^2$ is
given by \cite{Bollerslev87}:
\begin{eqnarray*}
\text{corr}(\epsilon_t^2, \epsilon^2_{t-n}) &=& \sum_{i=1} ^{p \vee q}
(\alpha_i + \beta_i) \rho_{n-i} \text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $\max(p, q)$.

\item {\bf Stochastic Volatility Models}\label{sec:SLV_model}

For the purpose of intraday returns, we specify the stochastic
volatility model as
\begin{eqnarray}
   r_{t, t-h} &=& \ln S_{t} - \ln S_{t-h} \nonumber \\
   r_{t, t-h} &=& \mu + \sigma_{t, t-h} b_t \label{eq:SLV_spec}
\end{eqnarray}
where $S_t$ is the price of the asset at time $t$; $b_t \sim N(0,
1)$; $r_{t, t-h}$ is the return over the time interval $[t-h, t]$.
For simplicity, in the rest of this chapter we shall just write
$t$ for the subscript ``t, t-h'', since the time interval $h$ is fixed
for each time series and is a known constant.

Andersen and Bollerslev et al proved the following in
\cite{Andersen03} (theorem 2):
\begin{equation}
  \label{eq:normal_r}
  r_t|\mathcal{F}_{t-h} \sim N(\int_{0}^h \mu_{t-h+s} ds, \int_{0}^h
  \sigma_{t-h+s}^2 ds)
\end{equation}
In plain words, given all the information up to time $t-h$, the
distribution of $r_t$ is Gaussian with integrated mean and variance.
The variance of this conditional distribution, $\int_{0}^h
\sigma_{t-h+s}^2 ds$, can be approximated by \cite{Protter05,
  Andersen03}:
\begin{equation}
  \label{eq:rv_def}
  \int_{0}^h \sigma_{t-h+s}^2 ds = \sum_{k=1}^n \left(
    S_{t-h+kh/n} - S_{t-h+(k-1)h/n} \right)^2
\end{equation}
where $n$ is a chosen constant. The square root of the right hand side
of equation \ref{eq:rv_def} is the realized volatility, call it
$\hat{\sigma}_t$.

@@

To compare GARCH and stochastic log-volatility models at the face of
intraday returns, we study the {\it Nordea Bank} 15-minute returns
during the period 2012/01/16 - 2012/04/20 in \S\ref{chp:nordea_15min},
and {\it Volvo B} 30-minute returns during 2013/10/10 - 2014/04/04 in
\S\ref{sec:volvo}. Another 5 intraday series are also studied and
collected in appendix \ref{chp:appendix2}.

\section{Case Study: Nordea 15-minute Returns}
\label{chp:nordea_15min}
In this section we investigate the Nordea 15-minute returns sampled
during the period 2012/01/16 - 2012/04/20. In total, these amount to
2022 returns. We use the first 80\% (1617) for model estimation and  the
remaining 20\% (405) for comparing with model forcasts. In section
\ref{sec:nordea_15min_garch} we study the series with a GARCH model 
and in section \ref{sec:nordea_15min_arima} we study it with a
stochastic volatility model.

\subsection{GARCH Model}\label{sec:nordea_15min_garch}
First of all, to determine that the volatilities are auto-correlated
and/or that the squared returns and the volatilities are correlated, and
hence a GARCH(p, q) model is appropriate for the return series under
investigation, we plot the auto-correlation function (ACF) of the
squared returns. This is shown in figure
\ref{fig:nordea_15min_acf}. If the aforementioned features are absent
from the series, the auto-correlations are expected to be Gaussian
distributed with mean 0 and variance 1/T and hence mostly reside
within the confidence bounds set by $\pm 2/\sqrt{T}$
\cite{Bollerslev86, Bollerslev87}.

Clearly this is not the case in figure \ref{fig:nordea_15min_acf} ---
the first 5 auto-correlations are rather significant. Moreover, figure
\ref{fig:nordea_15min_vlt_acf} shows even more clearly that the
conditional variances of the series are correlated. These observations
suggest a GARCH(p, q) model can be appropriate.
\begin{figure}[htb!]
  \centering
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=90 257 105
    220]{../pics/nordea_15min_acf.pdf}
    \label{fig:nordea_15min_acf}
  }
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=90 259 104
    218]{../pics/nordea_15min_vlt_acf.pdf}
    \label{fig:nordea_15min_vlt_acf}
  }
  \caption{\small \it \ref{fig:nordea_15min_acf}: Auto-correlations
    (ACF) of the squared returns; \ref{fig:nordea_15min_vlt_acf}:
    Auto-correlations (ACF) of squared realized volatilities
    ($\hat{\sigma}^2_t$). The blue lines are confident bounds set at
    $\pm 2/\sqrt{T}$.}
\end{figure}

Starting with a GARCH(1,1) model and taking advantage of the knowledge
that the log-volatility $\ln \sigma_t$ has seasonality $s=33$, we fit
to the return series a GARCH(33, 33) model, limiting to lags 1 and 33
for both ARCH and GARCH parameters.
\begin{eqnarray*}
  r_t &=& \mu + \epsilon_t \\
  \epsilon_t &=& \sigma_t z_t \\
  \sigma^2_t &=& \alpha_0 + \alpha_1 \epsilon^2_{t-1} + \alpha_s
  \epsilon^2_{t-s} + \beta_1 \sigma^2_{t-1} + \beta_s \sigma^2_{t-s}
\end{eqnarray*}
Via maximum likelihood estimation, parameter values listed in table
\ref{tab:nordea_15min_garch} are obtained.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|}
    \hline
    Parameter & $\alpha_0$ & $\alpha_1$ & $\alpha_s$ & $\beta_1$ &
    $\beta_s$ \\
    \hline
    Value & $4.7833 \times 10^{-7}$ & 0.1600 & 0.0667 & 0.6846 &
    0.0342 \\
    \hline
  \end{tabular}
  \caption{\small \it GARCH model parameters}
  \label{tab:nordea_15min_garch}
\end{table}

@@

The above seasonal moving average model has the following
autocovariance structure \cite{BoxJenkins94}:
\begin{eqnarray*}
  \gamma_0 &=& \sigma_y^2 (1 + \theta^2)(1 + \Theta^2) \\
  \gamma_1 &=& -\sigma_y^2\theta(1 + \Theta^2) \\
  \gamma_s &=& -\sigma_y^2\Theta(1 + \theta^2) \\
  \gamma_{s+1} &=& \gamma_{s-1}\;=\;\sigma_y^2\theta\Theta
\end{eqnarray*}
These equations together with the measured autocorrelations make
possible an initial estimate of the parameters $\theta$ and $\Theta$:
\begin{eqnarray*}
  {\varrho_{s+1}/\varrho_s} &=& {\gamma_{s+1}/\gamma_s} \;=\; -{\theta \over
    1 + \theta^2} \\
  {\varrho_{s+1}/\varrho_1} &=& {\gamma_{s+1}/\gamma_1} \;=\; -{\Theta \over
    1 + \Theta^2} \\
\end{eqnarray*}
Substituting in the measured values shown in table
\ref{tab:nordea_15min_w_acf},
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    $\varrho_1$ & $\varrho_{s-1}$ & $\varrho_s$ & $\varrho_{s+1}$ \\
    \hline
    -0.4703 &  0.2053 & -0.4564 &  0.2212 \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min $w_t$ autocorrelations}
  \label{tab:nordea_15min_w_acf}
\end{table}
we get
\begin{eqnarray*}
  \theta &=& 0.6890 \\
  \Theta &=& 0.6378
\end{eqnarray*}
Among the two roots of each of the 2nd order equations in the above,
we have chosen the one in the range $(-1, 1)$ so as to ensure
invertibility of the model \cite{BoxJenkins94}.

@@

The first 4 moments of the Johnson Su distribution are expressible
in closed form in $\gamma, \delta, \lambda, \xi$ \cite{Shang2004}.
% \begin{eqnarray*}
%   w &=& \exp{1 \over \delta^2} \\
%   \Omega &=& {\gamma \over \delta} \\
%   \text{E}(y) &=& -w^{1/2} \lambda \sinh\Omega + \xi\\
%   \text{std}(y) &=& \lambda \left[{1 \over 2}(w-1)(w\cosh 2\Omega +
%     1)\right]^{1/2} \\
%   \text{skewness}(y) &=& {
%     \sqrt{(1/2)w(w-1)} [w(w+2)\sinh 3\Omega + 3\sinh\Omega]
%     \over
%     (w\cosh 2\Omega + 1)^{3/2}} \\
%   \text{kurtosis}(y) &=& {
%     w^2(w^4 + 2w^3 + 3w^2 - 3)\cosh 4\Omega + 4w^2 (w+2) \cosh 2\Omega
%     + 3(2w+1) \over
%     2(w\cosh 2\Omega + 1)^2 }
% \end{eqnarray*}
By matching the theoretical expressions of the moments with their
measured values, and taking help from published tables
\cite{Johnson1965}, one can solve for the parameters $\gamma, \delta,
\lambda, \xi$.

@@

% The eventual MLE is done in Matlab with the ``active set''
% algorithm. The initial as well as the final estimation results are
% listed in table \ref{tab:nordea_15min_js_param}:
% \begin{table}[htb!]
%   \centering
%   \begin{tabular}{|c|c|c|c|c|c|c|}
%     \hline
%     & $\gamma$ & $\delta$ & $\lambda$ & $\xi$ & $\theta$ & $\Theta$ \\
%     \hline
%     initial estimate & 0.1476 & 1.5121 & 0.3633 & 0.0454 & 0.6890 &
%     0.6378 \\
%     \hline
%     MLE estimate & 0.1319 & 1.5266 & 0.3735 & 0.0410 & 0.6639 & 0.6025
%     \\
%     \hline
%   \end{tabular}
%   \caption{\small \it Nordea 15min estimation results}
%   \label{tab:nordea_15min_js_param}
% \end{table}

@@

\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.55, clip=true, trim=27 287 0
    154]{../pics/nordea_2012.pdf}
  \caption{\small \it Blue: SV forecasts; Green: GARCH forecasts; Red:
    sample mean forecasts. Left: $P(\ln \sigma^F_t - \ln \hat{\sigma}_t < x)$;
    Right: $P(\ln \sigma^F_t - \ln \hat{\sigma}_t > x)$. Horizontal: $x$.}
  \label{fig:nordea_2012}
\end{figure}

Another measure of the forecasts' quality can be the percentage of good
forecasts, where the criterion of ``good'' is defined, respectively,
as deviating no more than 1\%, 5\%, or 10\% from the measured value of
the realized volatility. Table \ref{tab:nordea_2012_good} shows the
respective percentage of the 3 kinds of forecasts. Again in this table
it is seen that the SV model gives more accurate forecasts than does
GARCH.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${\ln \sigma^F_t - \ln \hat{\sigma}_t \over |\ln \hat{\sigma}_t|}$
    & SV & GARCH & sample mean \\
    \hline
    1\% & 21.73\% & 13.83\% & 11.36\% \\
    \hline
    5\% & 74.57\% & 70.62\% & 53.33\% \\
    \hline
    10\% & 97.28\% & 93.83\% & 89.14\% \\
    \hline
  \end{tabular}
  \caption{\small \it The percentage of ``good'' forecasts when the
    criterion of being good is deviating no more than 1\%, 5\% or
    10\%.}
  \label{tab:nordea_2012_good}
\end{table}

\section{Case study: Volvo 30-minute Returns}
\label{sec:volvo_30}
In this section we model the log-volatility of {\it Volvo B} 30-minute
returns during the period 2013/10/10 - 2014/04/04. This series
contains 1884 log-volatilities computed using 2-minute returns in each
30-minute interval. Among them we use the first 1507 for model
estimation and the last 377 for forecast and model verification.

The left plot of figure \ref{fig:volvo_inno_and_lv_acf} shows the
distribution of $(r_t - \E(r_t))/\sigma_t$. We observe in the figure a
nice Gaussian variate, so we can be sure that the sum of squared
2-minute returns makes a good approximation to the variance of
30-minute returns in this particular case.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.4, clip=true, trim=0 256 0
    151]{../pics/volvo_inno_and_lv_acf.pdf}
    \label{fig:volvo_inno_and_lv_acf}
  \caption{\small \it Left: probability plot of $(r_t -
    \E(r_t))/\sigma_t$; Right: auto-correlations of $\ln \sigma_t$}
\end{figure}

@@

\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=10 282 0
    243]{../pics/volvo_slv_garch_cmp.pdf}
    \label{fig:volvo_slv_garch_cmp}
  \caption{\small \it Blue: SV forecasts; Green: GARCH forecasts; Red:
    sample mean forecasts. Left: $P(\ln \sigma^F_t - \ln
    \hat{\sigma}_t < x)$; Right: $P(\ln \sigma^F_t - \ln
    \hat{\sigma}_t > x)$. Horizontal: $x$.}
\end{figure}
Figure \ref{fig:volvo_slv_garch_cmp} shows, as in the previous case of
Nordea Bank 15-minute returns, the SV model performs the best, GARCH
the second, and the sample mean the worst. However, when it comes to
over-estimates, the sample mean appears to be the best estimator,
while SV excels over GARCH. But a check of $\E(\ln\sigma_t)$ over
the sample for estimation (1507 data points) and over the sample for
comparison (377 data points) reveals that the first sample has mean
-6.3127 while the second has -6.2072. This increment explains the low
probability of over-estimation when using the first sample mean as
forecast.

Table \ref{tab:volvo_good_percentage} compares the fraction of
``good'' estimates as measured by ${|\ln\sigma^F_t -
  \ln\hat{\sigma}_t| \over |\ln \hat{\sigma}_t|}$ being less than 1\%,
5\% and 10\%.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ &
    SV & GARCH & Sample Mean \\
    \hline
    1\% & 21.75\% & 12.20\% & 14.06\% \\
    \hline
    5\% & 71.62\% & 61.54\% & 66.31\% \\
    \hline
    10\% & 92.31\% & 88.06\% & 89.92\% \\
    \hline
  \end{tabular}
  \caption{\small \it Fraction of ``good'' forecasts as defined by
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ being less than 1\%, 5\% and 10\%.}
  \label{tab:volvo_good_percentage}
\end{table}
We see from the table that the SV model consistently excels over
the other two alternatives. In addition, it is also noted that the
GARCH forecast is even worse than the sample mean. This is surprising
but not totally unexpected --- with only 3 parameters, the GARCH(1,1)
model can only describe the most prominent auto-correlations in the
volatility. When the volatility is influenced by relatively weak
auto-correlations at several different time lags, the GARCH forecast
cannot be expected to have good accuracy.

In this particular case, we see that the ARIMA model has 3 relatively
small moving average coefficients, located at lags 2, 3, and 4 and
evaluated to 0.06, 0.06, 0.03, suggesting a scattered auto-correlation
structure, so the GARCH model cannot be expected to perform very
well. In contrast, the Volvo 15-minute returns studied in
\S\ref{sec:volvo} has more concentrated auto-correlations --- 0.12 and
0.06 at lags 2 and 3 --- thus the GARCH(1,1) model is also found to
perform better and even marginally better than the SV model for the
same series.
